%\pdfoptionpdfminorversion 6
%%& --> Document class <--
\documentclass[
    final,
% 	draft,
    paper=a4,
    pagesize=auto,
    fontsize=10pt, % original 11pt
    version=last,
]{scrbook}

% To use the upper directory for sources
\makeatletter
\def\input@path{{./literature/}}
\makeatother

\usepackage[usenames,dvipsnames]{color}
\usepackage{hyperref}
\hypersetup{
    breaklinks,
    colorlinks,
    linkcolor=gray,
    citecolor=gray,
    urlcolor=gray,
    pdftitle={The Tensor Network Approach to Efficient and Explainable AI},
    pdfauthor={Alex Goessmann}
}

\usepackage{tikz}
\usepackage{graphicx}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}

\usepackage{listings}
\usepackage{verbatim}
\usepackage{etoolbox}
\usepackage{braket}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{bbm}
\usepackage{bm}
\usepackage{algpseudocode}
\usepackage{algorithm}

% Bibliography
\DeclareUnicodeCharacter{FB01}{fi}
\usepackage[backend=biber, style=alphabetic]{biblatex}

\input{./macros/general_macros.tex}
\input{./macros/tc_macros.tex}
\input{./macros/tikz_macros.tex}

\input{./scrbook_compilation/preamble.tex}
\input{./scrbook_compilation/layout-macros.tex}

\pretolerance=500
\tolerance=100
\emergencystretch=10pt

\newcommand{\sect}[1]{\section{#1}}
\newcommand{\subsect}[1]{\subsection{#1}}
\newcommand{\subsubsect}[1]{\subsubsection{#1}}

\newcommand{\red}[1]{#1}

\begin{document}

    \listoftodos

%%% --> Front part <--
    \frontmatter

% title page
    \input{./OtherContent/title.tex}

    % define pagestyle
    \pagestyle{scrplain}

    % table of contents
    \cleardoublepage

    \pdfbookmark[chapter,color=blue,bold=true]{\contentsname}{toc}
    \tableofcontents

    \newpage
    \textbf{Abstract} \\
    \input{./OtherContent/abstract.tex}

    \input{./OtherContent/introduction.tex}
    \input{./OtherContent/notation.tex}
    \input{./OtherContent/summary}

    \mainmatter

    \part{\partonetext}\label{par:one}

    The computational automation of reasoning is rooted both in the probabilistic and the logical reasoning tradition.
    Both draw on the same ontological commitment that systems have a factored structure, that is their states are described by assignments to a set of variables.
    Based on this commitment both approaches bear a natural tensor representation of their states and a formalism of the respective reasoning algorithms based on multilinear methods.

    \input{./PartI/probability_representation.tex}
    \input{./PartI/probability_reasoning.tex}

    \input{./PartI/logic_representation.tex}
    \input{./PartI/logic_reasoning.tex}


    \part{\parttwotext}\label{par:two}

    We now employ tensor networks to define architectures and algorithms for neuro-symbolic reasoning based on the logical and probabilistic foundations.
    Markov Logic Networks will be taken as generative models to be learned from data, using formula selecting tensor networks and likelihood optimization algorithms.

    \input{./PartII/formula_selection.tex}

    \input{./PartII/network_representation.tex}
    \input{./PartII/network_reasoning.tex}

    \input{./PartII/concentration.tex}

    \input{./PartII/fol_models.tex}


    \part{\partthreetext}\label{par:three}

    Based on the logical interpretation we often handle tensor calculus with specific tensors.
    Often, they are binary (that is their coordinates are in $\{0,1\}$ corresponding with a Boolean), and sparse (that is having a decomposition with less storage demand).
    We investigate it in this part in more depth the properties of such tensors, which where exploited in the previous parts.

    \input{./PartIII/coordinate_calculus.tex}
    \input{./PartIII/basis_calculus.tex}

    \input{./PartIII/sparse_calculus.tex}

    \input{./PartIII/message_passing.tex}

    \input{./PartIII/approximation.tex}

    \appendix
    \input{./OtherContent/implementation.tex}

    \begin{refcontext}[sorting=nyt]
	    \printbibliography
    \end{refcontext}

\end{document}